Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-15 manuscript/draft.Rmd
minor

table_caption = "A summary of the six models evaluated here. The time series
models were trained site-by-site, without access to any environmental
information. The environmental models were trained at a continental scale,
without information regarding which transects occurred at which site or during
which year. Most of the models were trained to predict richness directly.
However, the random forest SDMs were each trained to predict the occurrence of
individual species at each site, and these species-level predictions were added
up to produce richness estimates. The mistnet JSDM was trained to predict the
full species composition at each site, and the number of species in its
predictions was used as an estimate of richness."

@@

table_data %>% 
  knitr::kable(
    format = "latex", 
    escape = FALSE, 
    booktabs = TRUE,
    caption = table_caption
  ) %>% 
  add_header_above(c(" " = 2, "Predictors" = 3)) %>% 
  group_rows("Time series models", 1, 3) %>% 
  group_rows("Environmental models", 4, 6) %>% 
  kable_styling(position = "center")
```

@@

![Example predictions from our six models for a single site. The data points
from 1982 through 2003, connected by solid lines, were used for training all six
models; the remaining points were used for evaluating the models' forecasts. In
each panel, point estimates for each year are shown with lines; the darker
ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
indicates 95% confidence intervals. **A.** The time series models models were
trained using one site at a time, based solely on observed richness values
shown. The first two models ("average" and "naive") served as our simple
forecasting baselines. **B.** The environmental models were trained to predict
richness based on elevation, climate, and NDVI; unlike the time series models,
the environmental models' predictions change from year to year as environmental
conditions change.](../figures/model_predictions.png)

@@

@@

**[[xxx]].** We used observed richness values from 1982 (the first complete year
of NDVI data) to 2003 for training the models, and data from 2004 to 2013 was
used for testing their performance. We only used BBS routes from the continental
United States (i.e. routes where PRISM data was available), and restricted the
analysis to routes that were sampled during 70% of the years in the training
period (i.e., routes with at least 16 annual observations). The resulting
dataset included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique
sites, and included `r numbers$N_species` species. Site-level richness varied
from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
an average richness of `r numbers$richness_summary$mean` species.

@@

[[Note: we sometimes refer to the baselines as "time series models", but (1)
we're also using that name for just auto.arima and (2) the "average" model
doesn't use any time series.  Other possible names: site-level models (as
above); richness-only models; environment-free models...]]

@@

In contrast to the site-level time-series models, most attempts to predict
species richness focus on using correlative models based on environmental
variables. We tested three common variants of this approach: direct modeling of
species richness; stacking individual species distribution models; and joint
species distribution models (JSDMs). Following the standard approach, site-level
random effects were not included in these models as predictors, meaning that
this approach implicitly assumes that two sites with identical Bioclim,
elevation, and NDVI values should have identical richness distributions. As
above, we included observer effects and the associated uncertainty by running
these models 500 times (once per MCMC sample).

@@

![Performance of the six models we evaluated when predicting one year into the
future (2004) or ten years into the future (2013).  In general, the time series
models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
the predictions generally declined as the forecast horizon was
extended.](../figures/scatter.png)

@@

![Each model's squared error can be divided into two parts. If a model
consistently over-estimates or under-estimates the richness at a given site,
this contributes to its error in predicting the site-level mean. The second
component is based on errors in predicting fluctuations in a site's richness
over time. Both components of the mean squared error were lower for the time
series models than for the environmental models.](../figures/barcharts.png)

@@

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness will continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales (e.g. 10-30 years in this study), which would make
simpler models with no directional change more appropriate. Recent suggestions
that local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@dornelas2014; @vellend2013]. This lack of change in richness may be expected
even in the presence of substantial changes in environmental conditions and
species composition at a site due to the replacement of species from the
regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
sites used in this study show little change in richness [site-level SD of `r
numbers$resid_sd` species, after controlling for differences among observers;
see also @lasorte2005]. The absence of rapid change in this dataset is
beneficial for the absolute accuracy of forecasts across different sites: when a
past year's richness is already known, it is easy to estimate a future richness.
The site-level stability of the BBS data also explains why stacked SDMs perform
relatively well at predicting future richness, despite failing to capture
changes in richness over time. However, this stability also makes it difficult
to improve forecasts relative to simple baselines, since those baselines are
already close to representing what is actually occurring in the system. These
results suggest that simple time-series models and baselines should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should also focus on species composition, since lower-level
processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
contain more useful information [@harris2015].

**Observer effects.** In addition to consideration of the different process
models used for forecasting it is important to consider the observation models.
When working with any ecological dataset, there are imperfections in the
sampling process that have the potential to influence results. With large scale
surveys and citizen science datasets, such as the Breeding Bird Survey, these
issues are potentially magnified by the large number of different observers and
by major differences in the habitats and species being surveyed [@sauer1994].
Accounting for differences in observers reduced the average error in our point
estimates and also improved the coverage of the confidence intervals. In
addition, controlling for observer effects resulted in changes in which models
performed best, most notably reducing the relative performance of the naive
model's point estimates. This suggests that, prior to accounting for observer
effects, the naive model performed well because it was capable of accommodating
rapid shifts in estimated richness introduced by changes in the observer. These
kinds of rapid changes were difficult for the other time-series models to
accommodate, and so the average performance of the ARIMA and average models
improved once this source of observation error was addressed. This demonstrates
that modeling observation error can be important for properly estimating and
reducing uncertainty in forecasts and can also lead to changes in the best
methods for forecasting. We did not address differences in detection probability
across species and sites [@boulinier1998] since there is no clear way to address
this issue without making strong assumptions about the data (i.e., assuming
there is no biological variation in stops along a route), but this would be a
valuable addition to future forecasting models.

**Environmental uncertainty.** Future biodiversity forecasting efforts will also
need to begin to address the additional uncertainty introduced by the error
inherent in the process of forecasting the environmental conditions that
ecologists use as predictor variables. In this, and other hindcasting studies,
the environmental conditions for the "future" are known because the data has
already been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts.
Thus, it is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017], as we did with the
uncertainty in observer-level effects. Difficulty in forecasting future
environmental conditions at small scales will present continued challenges for
models incorporating environmental variables, and this may result in a continued
advantage for simple time-series based approaches.

@@

Evalutating a models predictive performance across time is critical. Hindcasting
uses a temporal out of sample validation to mimic how well a model would have
performed had it been run in the past. For example using occurance data from the
early 20th century to model distributions which are validated with late
20th century occurances. Dense timeseries, such as yearly observations, are
desirable to also evalulate the forecast horizon (see #6), but this is not
a strict requirement.

>>>>

table_caption = "A summary of the six models evaluated here. The single-site
models were trained site-by-site, without access to any environmental
information. The environmental models were trained at a continental scale,
without information regarding which transects occurred at which site or during
which year. Most of the models were trained to predict richness directly.
However, the random forest SDMs were each trained to predict the occurrence of
individual species at each site, and these species-level predictions were added
up to produce richness estimates. The mistnet JSDM was trained to predict the
full species composition at each site, and the number of species in its
predictions was used as an estimate of richness."

@@

table_data %>% 
  knitr::kable(
    format = "latex", 
    escape = FALSE, 
    booktabs = TRUE,
    caption = table_caption
  ) %>% 
  add_header_above(c(" " = 2, "Predictors" = 3)) %>% 
  group_rows("Single-site models", 1, 3) %>% 
  group_rows("Environmental models", 4, 6) %>% 
  kable_styling(position = "center")
```

@@

![Example predictions from our six models for a single site. The data points
from 1982 through 2003, connected by solid lines, were used for training all six
models; the remaining points were used for evaluating the models' forecasts. In
each panel, point estimates for each year are shown with lines; the darker
ribbon indicates 1 standard deviation of uncertainty, and the lighter ribbon
indicates 95% confidence intervals. **A.** The single-site models models were
trained using one site at a time, based solely on observed richness values
shown. The first two models ("average" and "naive") served as our simple
forecasting baselines. **B.** The environmental models were trained to predict
richness based on elevation, climate, and NDVI; unlike the single-site models,
the environmental models' predictions change from year to year as environmental
conditions change.](../figures/model_predictions.png)

@@

We used observed richness values from 1982 (the first year of complete
environmental data) to 2003 for training the models, and data from 2004 to 2013
was used for testing their performance. We only used BBS routes from the
continental United States (i.e. routes where climate data was available @prism),
and we restricted the analysis to routes that were sampled during 70% of the
years in the training period (i.e., routes with at least 16 annual
observations). The resulting dataset included `r numbers$N_runs` transect runs
of `r numbers$N_sites` unique sites, and included `r numbers$N_species` species.
Site-level richness varied from `r numbers$richness_summary$min` to `r
numbers$richness_summary$max` with an average richness of `r
numbers$richness_summary$mean` species.

@@

@@

@@

In contrast to the single-site models, most attempts to predict
species richness focus on using correlative models based on environmental
variables. We tested three common variants of this approach: direct modeling of
species richness; stacking individual species distribution models; and joint
species distribution models (JSDMs). Following the standard approach, site-level
random effects were not included in these models as predictors, meaning that
this approach implicitly assumes that two sites with identical Bioclim,
elevation, and NDVI values should have identical richness distributions. As
above, we included observer effects and the associated uncertainty by running
these models 500 times (once per MCMC sample).

@@

![Performance of the six models we evaluated when predicting one year into the
future (2004) or ten years into the future (2013).  In general, the single-site
models (**A.**) outperformed the environmental models. (**B.**) The accuracy of
the predictions generally declined as the forecast horizon was
extended.](../figures/scatter.png)

@@

![Each model's squared error can be divided into two parts. If a model
consistently over-estimates or under-estimates the richness at a given site,
this contributes to its error in predicting the site-level mean. The second
component is based on errors in predicting fluctuations in a site's richness
over time. Both components of the mean squared error were lower for the
single-site models than for the environmental models.](../figures/barcharts.png)

@@

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness will continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales (e.g. 10-30 years in this study), which would make
simpler models with no directional change more appropriate. Recent suggestions
that local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@dornelas2014; @vellend2013]. This lack of change in richness may be expected
even in the presence of substantial changes in environmental conditions and
species composition at a site due to the replacement of species from the
regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
sites used in this study show little change in richness [site-level SD of `r numbers$resid_sd` species, after controlling for differences among observers;
see also @lasorte2005]. The absence of rapid change in this dataset is
beneficial for the absolute accuracy of forecasts across different sites: when a
past year's richness is already known, it is easy to estimate a future richness.
The site-level stability of the BBS data also explains why stacked SDMs perform
relatively well at predicting future richness, despite failing to capture
changes in richness over time. However, this stability also makes it difficult
to improve forecasts relative to simple baselines, since those baselines are
already close to representing what is actually occurring in the system. These
results suggest that single-site models should be actively considered for
forecasts of richness and other stable aspects of biodiversity. Our results also
suggest that future efforts to understand and forecast biodiversity should also
focus on species composition, since lower-level processes are expected to be
more dynamic [@ernest2001; @dornelas2014] and contain more useful information
[@harris2015].

**Observer effects.** In addition to consideration of the different process
models used for forecasting it is important to consider the observation models.
When working with any ecological dataset, there are imperfections in the
sampling process that have the potential to influence results. With large scale
surveys and citizen science datasets, such as the Breeding Bird Survey, these
issues are potentially magnified by the large number of different observers and
by major differences in the habitats and species being surveyed [@sauer1994].
Accounting for differences in observers reduced the average error in our point
estimates and also improved the coverage of the confidence intervals. In
addition, controlling for observer effects resulted in changes in which models
performed best, most notably reducing the relative performance of the naive
model's point estimates. This suggests that, prior to accounting for observer
effects, the naive model performed well because it was capable of accommodating
rapid shifts in estimated richness introduced by changes in the observer. These
kinds of rapid changes were difficult for the other single-site models to
accommodate, and so the average performance of the ARIMA and average models
improved once this source of observation error was addressed. This demonstrates
that modeling observation error can be important for properly estimating and
reducing uncertainty in forecasts and can also lead to changes in the best
methods for forecasting. We did not address differences in detection probability
across species and sites [@boulinier1998] since there is no clear way to address
this issue without making strong assumptions about the data (i.e., assuming
there is no biological variation in stops along a route), but this would be a
valuable addition to future forecasting models.

**Environmental uncertainty.** Future biodiversity forecasting efforts will also
need to begin to address the additional uncertainty introduced by the error
inherent in the process of forecasting the environmental conditions that
ecologists use as predictor variables. In this, and other hindcasting studies,
the environmental conditions for the "future" are known because the data has
already been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts.
Thus, it is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017], as we did with the
uncertainty in observer-level effects. Difficulty in forecasting future
environmental conditions at small scales will present continued challenges for
models incorporating environmental variables, and this may result in a continued
advantage for simple single-site approaches.

@@

Evalutating a models predictive performance across time is critical. Hindcasting
uses a temporal out of sample validation to mimic how well a model would have
performed had it been run in the past. For example using occurance data from the
early 20th century to model distributions which are validated with late
20th century occurances. Dense time series, such as yearly observations, are
desirable to also evalulate the forecast horizon (see #6), but this is not
a strict requirement.
